Palaeodistribution of crop progenitors and wild foods in the Southern Levant

### 2023-08-04 analysis/analysis.Rmd
Model taxon-by-taxon

taxa <- c("Hordeum spontaneum" = "Hordeum spontaneum")
          # "Bolboshoenus spp." = "Bolboschoenus")

@@

# Remove duplicated coordinates (as recommended by CITE)
occ |>
  group_by(taxon) |>
  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE) |>
  ungroup() ->
  occ

# Convert to sf
occ |>
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
           crs = 4326, remove = FALSE) ->
  occ
```

```{r plot-occ-distribution}
ggplot(occ) + 
  facet_wrap(vars(taxon)) +
  layer_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
  layer_spatial(occ, alpha = 0.1, symbol = 16) +
  layer_spatial(ne_countries, fill = NA) +
  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
           ylim = w_asia[c("ymin", "ymax")],
           crs = utm37n,
           default_crs = latlong)
```

@@

```{r psabsences, include=FALSE}
n_background <- 1000

occ |>
  mutate(present = TRUE) |>
  bind_rows(background_sample(s_levant_land, 
                              n_background,
                              coord_x = "decimalLongitude",
                              coord_y = "decimalLatitude", 
                              taxon = "Hordeum spontaneum", 
                              present = FALSE)) ->
  occ

# Weights
#occ %>% 
#  group_by(taxon, occurrence) %>% 
#  mutate(weight = 1 + (n_psab / n() - 1)) ->
#  occ
```

@@

```{r training-climate}
occ <- cbind(occ, st_extract(climate$cur, occ))
```

@@

```{r model}
# TODO: Where do the missing value(s) actually come from?
occ |>
  as_tibble() |>
  select(present, starts_with("bio")) |>
  drop_na() ->
  rf_data

rand_forest(mode = "regression") |>
  fit(as.numeric(present) ~ ., data = rf_data) ->
  rf_fit
```

@@

```{r predict}
rf_pred <- predict(climate$cur, rf_fit)
```

>>>>

taxa <- c("Triticum dicoccoides" = "Triticum dicoccoides",
          "Triticum urartu" = "Triticum urartu",
          "Hordeum spontaneum" = "Hordeum spontaneum",
          "Bolboshoenus spp." = "Bolboschoenus")

@@

# Remove redundant observations of the same taxon in the same place
occ |>
  group_by(taxon) |>
  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE) |>
  ungroup() ->
  occ

# Drop unneeded columns, normalise column names and convert to sf
occ |>
  select(
    taxon, gbif_key = key, genus, species,
    longitude = decimalLongitude, latitude = decimalLatitude
  ) |>
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = 4326, remove = FALSE) ->
  occ
```

```{r plot-occ-distribution}
ggplot(occ) + 
  facet_wrap(vars(taxon)) +
  layer_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
  layer_spatial(occ, alpha = 0.2) +
  layer_spatial(ne_countries, fill = NA) +
  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
           ylim = w_asia[c("ymin", "ymax")],
           crs = utm37n,
           default_crs = latlong)
```

@@

```{r data-background}
n_background <- 10000

occ |>
  mutate(present = TRUE) |>
  group_by(taxon) |>
  nest(.key = "occ_data") |>
  mutate(bg_data = map(1:n(), \(i) { 
    background_sample(
      region = s_levant_land, 
      N = n_background,
      coord_x = "longitude", coord_y = "latitude",
      present = FALSE
    )
  })) |>
  mutate(data = map2(occ_data, bg_data, bind_rows)) |>
  select(taxon, data) |>
  unnest(data) |>
  st_sf() ->
  occ
```

@@

```{r training-climate}
occ <- cbind(occ, st_extract(climate$cur, occ)) |> select(-geometry.1)
```

@@

There are many methods for ENM.
Random Forest is one of the best performing, providing it is appropriately parameterised to account for the class imbalance between presence and background samples (balanced random forest) [@ValaviEtAl2021a; @ValaviEtAl2021b].

We used the R package ranger [@WrightZiegler2017].

```{r model}
rf_pred_vars <- function() {
  c(any_of("present"), starts_with("bio"))
}

# Drop any rows with NA values in any of the predictor variables
# TODO: Where do these actually come from? There's usually only a handful.
occ <- drop_na(occ, rf_pred_vars())

# Model
# Parameters adapted from the example of a down-sampled RF model given in
#   https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
# TODO: tune?

rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger",
    probability = TRUE,
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1)
  ) -> 
  rf_model

# Fit to each taxon
occ |>
  nest(.by = taxon) |>
  mutate(
    data = map(data, as_tibble),
    fit = map(data, \(data) {
      fit(rf_model, factor(present) ~ ., data = select(data, rf_pred_vars()))
    })) ->
  rf_fits
```

@@

```{r predict}
rf_fit <- rf_fits$fit[[1]]

# TODO: predict on all periods (maybe combine stars and collapse into dimension first?)
rf_pred_cur <- predict(climate$cur, rf_fit, type = "prob")
rf_pred_eh <- predict(climate$EH, rf_fit, type = "prob")
rf_pred_yds <- predict(climate$YDS, rf_fit, type = "prob")
rf_pred_ba <- predict(climate$BA, rf_fit, type = "prob")
rf_pred_hs1 <- predict(climate$HS1, rf_fit, type = "prob")
rf_pred_lgm <- predict(climate$LGM, rf_fit, type = "prob")

rf_pred <- c(rf_pred_cur, rf_pred_eh, rf_pred_yds, rf_pred_ba, rf_pred_hs1,
             rf_pred_lgm)
names(rf_pred) <- c("cur", "EH", "YDS", "BA", "HS1", "LGM")

ggplot() + 
  geom_stars(data = st_redimension(st_crop(rf_pred, w_asia))) + 
  facet_wrap(~new_dim)
```
